---
title: "Finding Diseases with Shared Genetic Causes"
author: "Brian M Schilder and Shawn T O'Neil"
date: "Vignette updated: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i>"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finding Diseases with Shared Genetic Causes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include=FALSE, message=FALSE}
options(width = 300)

# Run live/networked chunks only when building pkgdown (or when explicitly forced).
live_eval <- identical(tolower(Sys.getenv("IN_PKGDOWN", "false")), "true") ||
    identical(tolower(Sys.getenv("MONARCHR_VIGNETTE_EVAL", "false")), "true")

knitr::opts_chunk$set(eval = live_eval, echo = TRUE, fig.width = 10, message = FALSE, warning = FALSE)
```

# Motivation

Genes and variants that cause one disease sometimes contribute to others as well,
a phenomenon known as pleiotropy. Phenome-wide association studies (PheWAS)
investigate this statistically, testing variants for association with many
phenotypes across large cohorts. Before (or alongside) such a study, it is useful
to ask what is already known: which other diseases share curated genetic causes
with a disease of interest?

The Monarch KG integrates curated gene- and variant-disease relationships from
sources such as ClinVar, OMIM, Orphanet, and ClinGen, along with the disease
hierarchy of the Mondo ontology. Here we use `monarchr` to start from a disease
(cardiomyopathy), collect its subtypes and their known genetic causes, find other
diseases attributed to the same variants, and summarize which genes link them.

Note that this is a knowledge-graph lookup, not a statistical association study:
results reflect curated, largely Mendelian, causal annotations, and are best
treated as a starting point for hypotheses about shared genetics rather than as
evidence of association. The full analysis takes a few minutes to run.

## 1. Get all subtypes of a disease

First we find the Mondo identifier for cardiomyopathy. A search returns several
related terms, so we pick the general one, `MONDO:0004994`, and fetch it:

```{r}
library(monarchr)
library(tidygraph)
library(dplyr)

monarch_search("cardiomyopathy", category = "biolink:Disease", limit = 5)

cardiomyopathy <- monarch_engine() |>
    fetch_nodes(query_ids = "MONDO:0004994")
```

Then we collect all of its subtypes by following `biolink:subclass_of` edges
inward, transitively:

```{r}
cm_subtypes <- cardiomyopathy |>
    expand(
        predicates = "biolink:subclass_of",
        direction = "in",
        transitive = TRUE
    )

cm_subtypes
```

## 2. Get genes and variants known to cause these diseases

Next we add genes and variants connected to any of these diseases by a causal or
genetic-association predicate. Most of the variants come from ClinVar:

```{r}
cm_genetics <- cm_subtypes |>
    expand(
        categories = c("biolink:Gene", "biolink:SequenceVariant"),
        predicates = c(
            "biolink:causes",
            "biolink:gene_associated_with_condition",
            "biolink:genetically_associated_with"
        )
    )

table(nodes(cm_genetics)$pcategory)

# where the causal edges come from
edges(cm_genetics) |>
    filter(predicate == "biolink:causes") |>
    count(primary_knowledge_source)
```

## 3. Find other diseases caused by the same variants

Starting from just the variants, we follow `biolink:causes` edges out to the
diseases they cause, and `biolink:is_sequence_variant_of` edges out to the genes
they belong to. Then we remove the cardiomyopathies we started with, leaving the
*other* diseases these variants are known to cause:

```{r}
shared_hits <- cm_genetics |>
    filter(pcategory == "biolink:SequenceVariant") |>
    expand(
        predicates = "biolink:causes",
        categories = "biolink:Disease",
        direction = "out"
    ) |>
    expand(
        predicates = "biolink:is_sequence_variant_of",
        categories = "biolink:Gene",
        direction = "out"
    ) |>
    filter(!id %in% nodes(cm_subtypes)$id)

table(nodes(shared_hits)$pcategory)
```

## 4. Summarize shared genes

With thousands of variant nodes, plotting this graph directly isn't very
informative. Instead we collapse each *gene ← variant → disease* path into a
single gene-disease link, counting the variants that support it. We first merge
the two graphs so that we have the variants' links to both the cardiomyopathies
and the other diseases:

```{r}
kg_all <- kg_join(cm_genetics, shared_hits)

variant_gene <- edges(kg_all) |>
    filter(predicate == "biolink:is_sequence_variant_of") |>
    select(variant = subject, gene = object)

variant_disease <- edges(kg_all) |>
    filter(predicate == "biolink:causes", subject %in% variant_gene$variant) |>
    select(variant = subject, disease = object)

gene_disease <- inner_join(variant_gene, variant_disease,
    by = "variant", relationship = "many-to-many"
) |>
    count(gene, disease, name = "n_variants") |>
    mutate(is_cardiomyopathy = disease %in% nodes(cm_subtypes)$id)
```

A gene is *shared* if its variants cause at least one cardiomyopathy and at least
one other disease. The genes shared with the most other diseases include
well-known pleiotropic genes such as *LMNA*, *TTN*, and the RAS-pathway genes
(*KRAS*, *HRAS*, *BRAF*, *PTPN11*):

```{r}
gene_names <- nodes(kg_all) |>
    select(gene = id, symbol = name)

shared_genes <- gene_disease |>
    group_by(gene) |>
    summarize(
        n_cardiomyopathies = sum(is_cardiomyopathy),
        n_other_diseases = sum(!is_cardiomyopathy),
        n_variants = sum(n_variants)
    ) |>
    filter(n_cardiomyopathies > 0, n_other_diseases > 0) |>
    left_join(gene_names, by = "gene") |>
    arrange(desc(n_other_diseases))

shared_genes
```

Finally, we build a small graph of just these genes and the diseases they link,
and plot it. Each gene-disease edge here is computed, standing in for one or more
variants:

```{r, fig.height = 9}
shared_edges <- gene_disease |>
    filter(gene %in% shared_genes$gene) |>
    transmute(
        subject = gene,
        predicate = "computed:shared_variant_cause",
        object = disease,
        primary_knowledge_source = "computed",
        n_variants
    )

shared_nodes <- nodes(kg_all) |>
    filter(id %in% c(shared_edges$subject, shared_edges$object)) |>
    mutate(group = case_when(
        pcategory == "biolink:Gene" ~ "Gene",
        id %in% nodes(cm_subtypes)$id ~ "Cardiomyopathy",
        TRUE ~ "Other disease"
    ))

shared_graph <- tbl_kgx(nodes = shared_nodes, edges = shared_edges)

plot(shared_graph, layout = "fr", node_color = group, edge_color = NULL)
```

Clusters in this plot correspond to familiar groups of genetic disease: the
RASopathies (Noonan syndrome and related disorders) around the RAS-pathway genes,
limb-girdle and other muscular dystrophies around sarcomere and dystrophin-complex
genes, and mitochondrial disorders around mitochondrially encoded genes.
