---
title: "Rollups and Transitivity"
author: "Shawn T O'Neil"
date: "Vignette updated: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i>"
output:
  rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Rollups and Transitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include=FALSE, message=FALSE}
options(width = 300)
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, fig.width = 10, message = FALSE, warning = FALSE)
```

First, let's load some required libraries, and instantiate a session-cached `monarch_engine()` for querying.

```{r}
library(monarchr)
library(tidygraph)
library(dplyr)

e <- monarch_engine()
```


Knowledge graphs frequently incorporate ontologies, which include complex hierarchies 
of classes and sub-classes. Let's visualize a couple of levels of this hierarchy for the phenotype `"leg phenotype"`
(removing `"leg phenotype"` as it will make the next examples clearer).


```{r}
phenos <- e |>
	fetch_nodes(name == "leg phenotype") |>
	expand_n(predicates = "biolink:subclass_of", 
					 categories = "biolink:PhenotypicFeature", 
					 direction = "out", 
					 n = 4) |>
	activate(nodes) |>
	filter(name != "leg phenotype")

plot(phenos)
```
It is not uncommon for data like this to come with additional information; if these were a set of disease diagnoses, 
we might have patient counts associated with each. Since patients receive diagnoses of varying specificity, 
there may be counts on any subtype.

Hypothesizing these phenotypes as diagnoses, we'll simulate some
count information, plotting it in the node labels:

```{r}
set.seed(42)

num_nodes <- nrow(nodes(phenos))
phenos_counted <- phenos |>
	activate(nodes) |>
	mutate(count = rpois(num_nodes, lambda = 5))

plot(phenos_counted, 
		 node_label = paste(name, 
		 									 " || count: ", count))

```

A "rollup" might thus ask, how many patients are associated with each phenotype,
if we include all of it's descendants? For example, "lower limb segment phenotype" (8 patients)
is a subclass of "limb segment phenotype" (4 patients), so the total number of "limb segment phenotype" 
patients includes both (12 patients). 

The `roll_up()` function allows us to compute this information. It is designed to work
with `dplyr`'s `mutate()` on node data: we provide the column specifying
information to aggregate, a function to apply over the values (amongst all descendants), 
and whether each node should include its own value in the aggregation.

```{r}
node_data <- nodes(phenos_counted) |> as.data.frame()
edge_data <- edges(phenos_counted) |> as.data.frame()
msg <- capture.output(print(node_data))
warning(paste(msg, collapse = "\n"))
msg <- capture.output(print(edge_data))
warning(paste(msg, collapse = "\n"))

# 
# phenos_counted_rolled <- phenos_counted |>
# 	activate(nodes) |>
# 	mutate(total = roll_up(count, fun = sum, include_self = TRUE))
# 
# plot(phenos_counted_rolled, 
# 		 node_label = paste0(name, 
# 		 									 " || count: ", count, 
# 		 									 " || total: ", total))

```

<!-- The corresponding **`roll_down()`** aggregates in the opposite direction (not shown). -->

<!-- ## Other aggregations, transferring information -->

<!-- When performing a rollup, each node receives the specified column, indexed to  -->
<!-- include only its descendants (and itself, if `include_self` is set). This is then passed to the  -->
<!-- aggregating function `fun`.  -->

<!-- To see how this can be useful, we'll start by introducing another function,  -->
<!-- `transfer()`. Much like `roll_up()`, this function is designed to be used with -->
<!-- `mutate()` on node data; its purpose is to transfer information across edges, -->
<!-- usually from nodes of one kind to another. We'll start by fetching all of the subtypes -->
<!-- of Niemann-Pick disease, and all known causal genes. -->


<!-- ```{r} -->
<!-- npc_genes <- e |> -->
<!-- 	fetch_nodes(name == "Niemann-Pick disease") |> -->
<!-- 	descendants() |> -->
<!-- 	expand(predicates = "biolink:causes") -->

<!-- plot(npc_genes) -->
<!-- ``` -->
<!-- Now, we might wish for disease nodes to have an attribute reflecting their causal genes. -->
<!-- This information is captured in the graph, but not as a part of the nodes. The `transfer()` -->
<!-- function 'pulls' information across edges: -->

<!-- ```{r} -->
<!-- npc_genes_causal <- npc_genes |> -->
<!-- 	activate(nodes) |> -->
<!-- 	mutate(caused_by = transfer(name, over = "biolink:causes", direction = "out")) -->

<!-- plot(npc_genes_causal, node_label = paste0(name,  -->
<!-- 																					 " || caused by: ", caused_by)) -->
<!-- ``` -->

<!-- Here, transfer is moving information 'over' (or across) `"biolink:causes"` -->
<!-- edges in an outward direction, along the direction of the edge. The transferred -->
<!-- information is being drawn from source nodes' `name`, resulting in a new `caused_by` -->
<!-- column in the node table.  -->

<!-- ```{r} -->
<!-- nodes(npc_genes_causal) |> -->
<!-- 	select(name, caused_by) -->
<!-- ``` -->

<!-- In cases where a transfer would result in multiple values being collected at the destination node,  -->
<!-- the result will be a list column. -->

<!-- To finish this example, we use `roll_up()` to collect, for each diseases, the set of genes -->
<!-- that cause it or any of its subtypes. -->

<!-- ```{r} -->
<!-- npc_genes_causal_rolled <- npc_genes_causal |> -->
<!-- 	activate(nodes) |> -->
<!-- 	mutate(any_caused_by = roll_up(caused_by,  -->
<!-- 																 fun = unique,  -->
<!-- 																 include_self = TRUE,  -->
<!-- 																 predicates = "biolink:subclass_of")) -->

<!-- plot(npc_genes_causal_rolled, node_label = paste0(name,  -->
<!-- 																					        " || any caused by: ", any_caused_by)) -->
<!-- ``` -->
<!-- The inclusion of `NA` values may not be desired (it signals that at least one -->
<!-- of the rolled nodes had a `caused_by` of `NA`). We could write an aggregating function -->
<!-- that removes `NA` and supply that; this would also a good use case for `purrr`'s  -->
<!-- `compose()` (`fun = compose(unique, na.omit)`). -->

<!-- ## Transitive closures and reductions -->

<!-- Let's return to the patient-count example, using the rolled-up data: -->

<!-- ```{r} -->
<!-- plot(phenos_counted_rolled,  -->
<!-- 		 node_label = paste0(name,  -->
<!-- 		 									 " || count: ", count,  -->
<!-- 		 									 " || total: ", total)) -->

<!-- ``` -->
<!-- It may be the case that to protect patient privacy (again, pretending these  -->
<!-- phenotypes are disease diagnoses associated with patients) we want to remove -->
<!-- nodes that have a `count` less than 6. If we do so however, we lose connectivity: -->

<!-- ```{r} -->
<!-- censored <- phenos_counted_rolled |> -->
<!-- 	activate(nodes) |> -->
<!-- 	filter(!count < 6) -->

<!-- plot(censored,  -->
<!-- 		 node_label = paste0(name,  -->
<!-- 		 									 " || count: ", count,  -->
<!-- 		 									 " || total: ", total)) -->

<!-- ``` -->
<!-- To fix this, we can first compute the `transitive_closure()` of the graph, with respect -->
<!-- to an edge predicate we want to treat as transitive (defaulting to `biolink:subclass_of`). We color edges -->
<!-- by `primary_knowledge_source` to highlight that newly created transitive edges are given knowledge source -->
<!-- `transitive_<predicate>`, but use the same predicate. The result is busy, and in general the number of transitive -->
<!-- edges can be $O(n^2)$ in the number of nodes. -->

<!-- ```{r} -->
<!-- phenos_closed <- phenos_counted_rolled |> -->
<!-- 	transitive_closure(predicate = "biolink:subclass_of") -->

<!-- plot(phenos_closed,  -->
<!-- 		 node_label = paste0(name,  -->
<!-- 		 									 " || count: ", count,  -->
<!-- 		 									 " || total: ", total), -->
<!-- 		 edge_color = primary_knowledge_source,  -->
<!-- 		 edge_linetype = predicate) -->

<!-- ``` -->
<!-- Now we can try our removal: -->

<!-- ```{r} -->
<!-- closed_censored <- phenos_closed |> -->
<!-- 	activate(nodes) |> -->
<!-- 	filter(!count < 6) -->

<!-- plot(closed_censored,  -->
<!-- 		 node_label = paste0(name,  -->
<!-- 		 									 " || count: ", count,  -->
<!-- 		 									 " || total: ", total), -->
<!-- 		 edge_color = primary_knowledge_source,  -->
<!-- 		 edge_linetype = predicate) -->
<!-- ``` -->

<!-- The graph retains its connectivity, but has redundant edges. The `transitive_reduction()` -->
<!-- function removes these, again according to a specified transitive predicate  -->
<!-- (defaulting again to `biolink:subclass_f`). -->

<!-- ```{r} -->
<!-- phenos_final <- closed_censored |> -->
<!-- 	transitive_reduction(predicate = "biolink:subclass_of") -->

<!-- plot(phenos_final,  -->
<!-- 		 node_label = paste0(name,  -->
<!-- 		 									 " || count: ", count,  -->
<!-- 		 									 " || total: ", total), -->
<!-- 		 edge_color = primary_knowledge_source,  -->
<!-- 		 edge_linetype = predicate) -->
<!-- ``` -->











